High geomagnetic field intensity in the late Mesoproterozoic recorded by Midcontinent Rift anorthosite xenoliths

### 2021-10-12 manuscript/Manuscript.tex
manuscript update

40 from a total of 86 anorthosite specimens and 9 out of a total of 69 diabase specimens passed our paleointensity selection criteria, yielding 7 anorthosite sites and 1 diabase sites with site-level paleointensity results that pass paleointensity selections. Example NRM/TRM plots (Arai plots; 
\cite{Arai1963a}) are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are shown in Table \ref{tab:PINT_result} and are plotted in Figure \ref{fig:PINT_cooling_corrected}. All measurement-level paleointensity experiment data can be accessed through the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

@@

The Beaver River anorthosite specimens yielded a high pass rate ($\sim$46.5\%) through our selection criteria. In distinct contrast, the Beaver River diabase specimens are all rejected by our selection criteria. Paleointensity experiment results from both lithologies show three types of dominant specimen behaviors on Arai plots (Figure \ref{fig:IZZI_examples}; \citeA{Arai1963a}): dominantly single-slope, double-slope, and zigzagging or sagging (Figure \ref{fig:IZZI_examples}). Note that almost all specimens have minimal secondary remanence component in thermal demagnetization results as shown in example orthogonal plots of all zero-field steps (Figure \ref{fig:IZZI_examples}). Typical specimen paleointensity behavior of the anorthosites display straight, single-slope Arai plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Figure \ref{fig:IZZI_examples}). We accept all specimen- and site-level absolute paleointensity estimates from the anorthosite xenoliths that pass the selection. In terms of the diabase specimens, all failed the selection because of double-slope behavior (fail FRAC selection), poor pTRM checks, and zigzagging (fail SCAT, DANG selection) Arai plots (Figure \ref{fig:IZZI_examples}). As is shown in Figure \ref{fig:IZZI_examples}, specimen BD2-5b has a high-temperature second slope with zigzagging behavior, poor pTRM checks, and low NRM/TRM ratio, in addition to the straight slope with steep NRM/TRM ratio over the lower temperature range. Although the BD2 specimens' lower temperature fraction of the Arai plots are straight andthe pTRM checks are consistent, lower temperature slopes from BD2 give consistently higher paleointensity estimates than those from site AX6, AX8, and AX10, which are anorthosite xenoliths included within BD2 (Figure \ref{fig:Geologic_map}; \citeA{Zhang2021a}). This higher estimate from the diabase lithology could be attributed to the systematic bias of using the lower temperature slope of the Arai plots demonstrated in Figure \ref{fig:IZZI_examples}. Thus we only accept the single-slope anorthosite specimens results for further analyses.

@@

The 20 mT alternating field treatment after in-field heating steps did not make any diabase specimen to pass the selection criteria. On the other hand, 27 out of 58 ($\sim$46.6\%) anorthosite specimens that had AF treatment passed selection and 13 out of 28 ($\sim$46.4\%) specimens without AF treatment passed selection. Paleointensity estimates from specimens with or without AF treatment in anorthosite AX10, AX11, AX12, AX13, AX16 do not show significant difference (Figure \ref{fig:PINT_cooling_corrected}). Specimen plots for AX11 particularly illustrates a case where all specimens within the site yield almost the same paleointensity estimate, regardless of the AF treatment. Therefore, the alternating field demagnetization might have improve the specimen behavior for those anorthosites that were subjected to this treatment but it did not result in significant increase in specimen success rate and did not result in significant difference for the site-level paleointensity estimates. Together with the narrow range of unblocking temperatures of their sister specimens \cite{Zhang2021b}, these data are consistent with the interpretation that the anorthosites with successful paleointensity results have a magnetic mineral assemblage that likely contain dominantly single domain, low Ti magnetite and they do not contain a significant population of MD or vortex state grains. Some anorthosite specimens and all diabase specimens still failed in result selections even though AF treatment was applied. It could be that they have a very high dominance of MD or vortex state grains with relatively moderate coercivity such that an alternating field with 20 mT peak field did not significantly mitigate their non-ideal paleointensity behaviors. Another cause for their failure could be the thermochemical alterations that happened during heating in air. As can be exampled by the zigzagging high temperature slope with poor pTRM checks in paleointensity plot for specimen BD2-5b (Figure \ref{fig:IZZI_examples}), the decrease in NRM/TRM ratios during higher temperature steps likely resulted from the formation of new ferromagnetic material within the specimen due to oxidation. Such alterations can allow specimens to acquire additional remanence during in-field steps that may not be fully removed by an AF treatment. Overall, the consistent specimen-level behavior makes the successful results from the anorthosites of high confidence.

@@

\caption{\footnotesize{Summary paleointensity results for specimens that passed our selection. Paleointensity results for specimens that passed quality criteria. B$_{anc}$ is the calculated ancient field intensity over the chosen temperature interval in $\mu$T. T$_{min}$ and T$_{max}$ indicate the temperature interval over which the best fit for paleointensity was defined. N is the number of steps used within the selected interval for paleointensity determination. FRAC is the fraction of remanence. NpTRM shows the number of pTRM checks within the selected interval for paleointensity determination. $\beta$ is the scatter parameter.  GAP-MAX is the maximum magnetization gap between two adjacent steps. MAD is the maximum angle of deviation. DANG is the deviation angle. SCAT is the scatter parameter. Paleolatitude is calculated from the inclination values reported in \citeA{Zhang2021b}. $\gamma$ is the gamma statistic that measures the angle between the last pTRM step used for paleointensity determination and the applied field direction. V(A)DM is the virtual (axial) dipole moment reported in $10^{21}$Am$^2$ ((ZAm$^2$)). }}

@@

\caption{\footnotesize{Summary statistics for the Q$_{PI}$ quality criteria of \citeA{Biggin2014a}.}}
\centering
\begin{tabular}{ccccccccccccc}
\hline
Site & N  & Age (Ma) & Method & AGE & STAT & TRM & ALT & MD & ACN & TECH & LITH & QPI \\ \hline
AX6  & 3  & 1091.8   & T+     & 1   & 0    & 1   & 1   & 1  & 1   & 0    & 0    & 5   \\
AX8  & 3  & 1091.8   & T+     & 1   & 0    & 1   & 1   & 1  & 1   & 0    & 0    & 5   \\
AX10 & 3  & 1091.8   & T+     & 1   & 0    & 1   & 1   & 1  & 1   & 0    & 0    & 5   \\
AX11 & 7  & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\
AX12 & 6  & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\
AX13 & 7  & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\
AX16 & 11 & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\ \hline
BD2  & 9  & 1091.7   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\ \hline
\end{tabular}
\label{tab:QPI}
\end{table}

@@

\subsection*{Late Mesoproterozoic geomagnetic field and the Precambrian observation records}

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{MCR_VADM.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual axial dipole moment (VADM) values from previous studies of the Midcontinent Rift rocks \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} and from the anorthosites that pass the selection criteria in this study. The median VADM value of all MCR rocks is higher than the median value of the past 300 myr. }}
\label{fig:MCR_VADM}
\end{figure}

considering \citeA{Nagy2021a} developed a new cooling rate correction model for uniaxial single domain magnetite and suggest that the model of \citeA{Halgedahl1980a} tend to over-correct for cooling rates and lead to underestimating ancient fields. If our absolute paleointensity values are an underestimating,

In addition, \cite{Cych2021a} developed a Bayesian probability method to estimate paleointensity - independent of the selection criteria, the Bayesian method fits for the curvature and generate probability distribution that guides credible interval for the paleoitnensity at a site level. The result distribution of Markov Chain Monte Carlo resampling are shown as violin plots in Figure \ref{}. The cooling rate corrected distributions show similar results as the frequentists' estimates based on the selection criteria derived results.

Address the spacial groups and using direction and intensity to address the issue of secular variation.

overall the PINT results are of high quality and the Qpi index average to be 6 to 7.

>>>>

40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity selection criteria, yielding 7 anorthosite sites and no diabase sites with site-level paleointensity results that pass paleointensity selections. Example NRM/TRM plots (Arai plots; 
\cite{Arai1963a}) are shown in Figure \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are shown in Table \ref{tab:PINT_result} and are plotted in Figure \ref{fig:PINT_cooling_corrected}. All measurement-level paleointensity experiment data can be accessed through the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

@@

The Beaver River anorthosite specimens yielded a high pass rate ($\sim$46.5\%) through our selection criteria. In distinct contrast, the Beaver River diabase specimens are all rejected by our selection criteria. Paleointensity experiment results from both lithologies show three types of dominant specimen behaviors on Arai plots (Figure \ref{fig:IZZI_examples}; \citeA{Arai1963a}): dominantly single-slope, double-slope, and zigzagging or sagging (Figure \ref{fig:IZZI_examples}). Note that almost all specimens have minimal secondary remanence component in thermal demagnetization results as shown in example orthogonal plots of all zero-field steps (Figure \ref{fig:IZZI_examples}). Typical specimen paleointensity behavior of the anorthosites display straight, single-slope Arai plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Figure \ref{fig:IZZI_examples}). We accept all specimen- and site-level absolute paleointensity estimates from the anorthosite xenoliths that pass the selection. In terms of the diabase specimens, all failed the selection because of double-slope behavior (fail FRAC selection), poor pTRM checks, and zigzagging (fail SCAT, DANG selection) Arai plots (Figure \ref{fig:IZZI_examples}). As is shown in Figure \ref{fig:IZZI_examples}, specimen BD2-5b has a high-temperature second slope with zigzagging behavior, poor pTRM checks, and low NRM/TRM ratio, in addition to the straight slope with steep NRM/TRM ratio over the lower temperature range. Although the BD2 specimens' lower temperature fraction of the Arai plots are straight andthe pTRM checks are consistent, lower temperature slopes from BD2 give consistently higher paleointensity estimates than those from site AX6, AX8, and AX10, which are anorthosite xenoliths included within BD2 (Figure \ref{fig:Geologic_map}; \citeA{Zhang2021a}). This higher estimate from the diabase lithology could be attributed to the systematic bias of using the lower temperature slope of the Arai plots demonstrated in Figure \ref{fig:IZZI_examples} and specimens from site BD2 all fail the criteria of having a maximum magnetizatiom gap between adjacent steps of 25\% of the total NRM. Thus we only accept the single-slope anorthosite specimens results for further analyses.

@@

The 20 mT alternating field treatment after in-field heating steps did not make any diabase specimen to pass the selection criteria. On the other hand, 27 out of 58 ($\sim$46.6\%) anorthosite specimens that had AF treatment passed selection and 13 out of 28 ($\sim$46.4\%) specimens without AF treatment passed selection. Paleointensity estimates from specimens with or without AF treatment in anorthosite AX10, AX11, AX12, AX13, AX16 do not show significant difference (Figure \ref{fig:PINT_cooling_corrected}). Specimen plots for AX11 particularly illustrates a case where all specimens within the site yield almost the same paleointensity estimate, regardless of the AF treatment. Therefore, the alternating field demagnetization might have improve the specimen behavior for those anorthosites that were subjected to this treatment but it did not result in significant increase in specimen success rate and did not result in significant difference for the site-level paleointensity estimates. Together with the narrow range of unblocking temperatures of their sister specimens \cite{Zhang2021b}, these data are consistent with the interpretation that the anorthosites with successful paleointensity results have a magnetic mineral assemblage that likely contain dominantly single domain, low Ti magnetite and they do not contain a significant population of MD or vortex state grains. Some anorthosite specimens and all diabase specimens still failed in result selections even though AF treatment was applied. It could be that they have a very high dominance of MD or vortex state grains with relatively moderate coercivity such that an alternating field with 20 mT peak field did not significantly mitigate their non-ideal paleointensity behaviors. Another cause for their failure could be the thermochemical alterations that happened during heating in air. As can be exampled by the zigzagging high temperature slope with poor pTRM checks in paleointensity plot for specimen BD2-5b (Figure \ref{fig:IZZI_examples}), the decrease in NRM/TRM ratios during higher temperature steps likely resulted from the formation of new ferromagnetic material within the specimen due to oxidation. Such alterations can allow specimens to acquire additional remanence during in-field steps that may not be fully removed by an AF treatment. Overall, the consistent specimen-level behavior makes the successful results from the anorthosites of high confidence. The paleointensity quality index (Q$_{PI}$) for the anorthosite sites are all above 3 and the summary statistics are shown in Table \ref{tab:QPI}.

@@

\caption{\footnotesize{Summary paleointensity results for specimens that passed our selection. Paleointensity results for specimens that passed quality criteria. B$_{anc}$ is the calculated ancient field intensity over the chosen temperature interval in $\mu$T. T$_{min}$ and T$_{max}$ indicate the temperature interval over which the best fit for paleointensity was defined. N is the number of steps used within the selected interval for paleointensity determination. FRAC is the fraction of remanence. NpTRM shows the number of pTRM checks within the selected interval for paleointensity determination. $\beta$ is the scatter parameter.  GAP-MAX is the maximum magnetization gap between two adjacent steps. MAD is the maximum angle of deviation. DANG is the deviation angle. SCAT is the scatter parameter. Paleolatitude is calculated from the inclination values reported in \citeA{Zhang2021b}. $\gamma$ is the gamma statistic that measures the angle between the last pTRM step used for paleointensity determination and the applied field direction. V(A)DM is the virtual (axial) dipole moment reported in $10^{21}$Am$^2$ (ZAm$^2$). }}

@@

\caption{\footnotesize{Summary statistics for the Q$_{PI}$ quality criteria of \citeA{Biggin2014a}.}}
\centering
\begin{tabular}{ccccccccccccc}
\hline
Site & N  & Age (Ma) & Method & AGE & STAT & TRM & ALT & MD & ACN & TECH & LITH & QPI \\ \hline
AX6  & 3  & 1091.8   & T+     & 1   & 0    & 1   & 1   & 1  & 1   & 0    & 0    & 5   \\
AX8  & 2  & 1091.8   & T+     & 1   & 0    & 1   & 1   & 1  & 1   & 0    & 0    & 5   \\
AX10 & 3  & 1091.8   & T+     & 1   & 0    & 1   & 1   & 1  & 1   & 0    & 0    & 5   \\
AX11 & 7  & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\
AX12 & 6  & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\
AX13 & 7  & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\
AX16 & 11 & 1091.8   & T+     & 1   & 1    & 1   & 1   & 1  & 1   & 0    & 0    & 6   \\ \hline
\end{tabular}
\label{tab:QPI}
\end{table}

@@

\subsection*{Late Mesoproterozoic geodynamo }
To best characterize the geomagnetic axial dipole field intensity during a certain time period requires that a paleointensity dataset to cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged and the geocentric axial dipole hypothesis (GAD) can be assumed. Thermal modeling results from \citeA{Zhang2021b} suggest that the anorthosite xenoliths were likely heated to tholeiitic magma temperature and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a few thousand years. In addition, the closely located sites AX12 and AX16 have very similar paleointensity estimates that are distinct from those of sites AX11 and AX13 which were emplaced $\sim$10 km away (Figure \ref{fig:Geologic_map}). The consistency within small regions but variations between regions could be the result of differential cooling history due to heterogeneity of the timing and depths of the Beaver River diabase intrusions, causing the anorthosites to capture paleosecular variations \textit{ca.} 1092 Ma. Overall, none of the anorthosite xenoliths record a V(A)DM value smaller than 40 ZAm$^2$ (Figure \ref{fig:PINT_cooling_corrected}).

Such a high estimate for the late Mesoproterozoic may invoke more careful examination of the geodynamo evolution through the Proterozoic and the timing for the inner core nucleation. Paleomagnetic observational records near the Ediacaran-Cambrian boundary have been interpreted to be consistent with a trend of decreasing axial dipole strength toward a very low value \cite{Lloyd2021a, Bono2019a}, followed by a rapid recovery (a few tens of million years) toward higher values soon after \cite{Thallner2021a}. Such a transition near this time can be consistent with the dynamo numerical simulations which adopt a rather high thermal conductivity of the core \cite{Davies2021a}. In particular, some models have suggested that the kick-start of rigorous compositional convection between the inner core and the outer core soon after the onset of inner core nucleation can result in almost an order of magnitude increase in Earth's axial dipole moment \cite{Davies2021a}.

Although more recent data

using paleomagnetic observational records is guided by the predictions from numerical simulations that a cooling liquid core with only thermal convection power to drive a dynamo would experience a progressive decay in its magnetic dipole field strength, and experience a rapid recover after the onset of inner core nucleation \cite{Davies2021a}. Previous studies has pointed to the Ediacaran-Cambrian boundary for such a weakening field behavior and recent studies has claim to found a rapid increase after the single crystal time which fits in the numerical models. However, such a recover may not be unique given the strong geodynamo estimate from the 1.1 Ga Midcontinent Rift. One could point to the increase from low values from the ca. 1140 Ma abitibi dikes to the ca. 1092 Ma anorthosites to be the result of recovery; another line of disagreement is between the Bangemall sills and the low field from Sept Ils. MCR being an anomaly? or the time scale we are looking at is too long?

Regardless these several lines of increase/decrease trend needs further explanations/considerations.

use a combined method of direction and intensity research for further examinations. Re-evaluation of the spatial and temporal variability captured by the existing Precambrian database. More data are needed for the Precambrian in order to better characterize the geodynamo evolution.

\begin{figure}[h!]
\noindent\includegraphics[width=\textwidth]{PINT_compilation.pdf}
\centering
\caption{\footnotesize{Compilation of calculated virtual (axial) dipole moment (VADM) values from the PINT database (v.2015.05; \url{http://earth.liv.ac.uk/ pint/}; \citeA{Biggin2015a}) with additional Mesoproterozoic data from \citeA{Sprain2018a, Kodama2019a}, as well as those from the Neoproterozoic \cite{Lloyd2021a, Lloyd2021b, Thallner2021a, Thallner2021b}. Paleointensity estimates from \citeA{Pesonen1983a} and \citeA{Kulakov2013a} are not included in the compilation due to the specimen-level double-slope behavior as discussed in the text. Overall the anorthosite xenoliths from this study record a high Mesoproterozoic field exceeding the value projected by the second order polynomial curve from \citeA{Bono2019a} which is based on an interpretation of there being a progressive decay of the geodynamo through the Proterozoic.}}
\label{fig:MCR_VADM}
\end{figure}
